An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-29 paper.tex
Update §5.3 introduction

%--------------------------------------------------------------------------------------------------
\subsection{Hierarchy and Communication Dynamics}
\label{sec:org-dyn:hierarchy}

@@

In the following, we quantitatively determine the effect that organisational
hierarchy levels have on communication patterns throughout the IETF. 
% We model communication as a network where a node represents an individual
% and a directed edge represents a communication sent from one individual to
% another~\cite{panzarasa2009patterns,viswanath2009evolution,klimt2004enron}.
% By repeating the analysis across time, we can build a full temporal
% network~\cite{masuda2016guide} of communication interactions between people
% in the system.In this study we consider only the network structure and the
% status at that time of the individuals, not the content of messages. 
We study how the email communication is impacted by the hierarchical roles
of individuals. According to the IETF mission statement~\cite{rfc3935}, the
IETF is made up of volunteers who collaborate to develop consensus-based
technical standards. This collaboration should be evident in their
communication practises.

@@

In the following we consider the following research questions: 
\begin{enumerate}
  \item
    Are higher levels of the hierarchy growing or becoming more
    centralised? Are they associated with an increased domination
    of the conversation?
    (\S\ref{sec:org-dyn:hierarchy:rq1})

@@

  \item
    How do people with differing roles communicate with each other?  Does
    information flow ``up" or ``down" the hierarchy?
    (\S\ref{sec:org-dyn:hierarchy:rq3})

@@

The first hypothesis focuses on the impact of the
``steepness"~\cite{anderson2010functions} of the hierarchy. This refers to
the fewer people in higher hierarchy levels the more steep the hierarchy
is, i.e a centralised control of decision making in higher levels. The
opposite of this is a ``diffused"~\cite{hussain2019voice} hierarchy where
the large amount of people in higher levels stifle lower level voice by
creating ``bystanders" who find it hard to speak up. In general, the
conclusions about steep hierarchies are confused, where some papers
conclude a positive and some a negative effect of steep hierarchies. We
explore what this means for the IETF in RQ1.

The second hypothesis regards the ``power
distance"~\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
individuals, which is the number of hierarchy levels between people who are
communicating. The focus is on the impact this has on people's voice or
lack thereof when the distance is large versus small. The conclusions from
this research area usually suggest that even small power distances have a
large effect of reducing the voice of the lower of the two levels. We
tackle this for the IETF in three different ways outlined in RQ2-4.

The IETF being a voluntary organisation gives a new perspective on these
hypotheses, as the existing literature focuses on commercial organisations.
In contrast to the above conclusions, we demonstrate that despite the
hierarchy becoming more ``diffused", participants in the higher levels in
the IETF hierarchy perform a ``facilitator" role, promoting discussion with
those they interact with and receiving a benefit themselves as a result. We
also show that higher level participants tend to be a focus of discussion
with remarks directed ``upward" toward them whereas Regular Participants
(RPs) engage more in group-discussion with less focus on an individual.
This suggests that the IETF does not conform to the notion that power
distance has a negative effect on communication, nor that a more diffused
hierarchy reduces the voice of lower levels.

>>>>

%--------------------------------------------------------------------------------------------------
\subsection{Hierarchy and Communication Patterns}
\label{sec:org-dyn:hierarchy}

@@

As a voluntary and consensus-driven \cite{RFC7282} standards development
organisation, one might expect the communication patterns in the IETF to
differ from those in businesses following a more traditional management
structure. Those in IETF leadership rules, whether working group chairs
or area directors, must lead by persuasion and example and have little in
the way of coercive power compared to a typical manager and this can be
expected to impact how they communicate.

In the following, we quantitatively determine the effect that organisational
hierarchy levels have on communication patterns throughout the IETF. 
% We model communication as a network where a node represents an individual
% and a directed edge represents a communication sent from one individual to
% another~\cite{panzarasa2009patterns,viswanath2009evolution,klimt2004enron}.
% By repeating the analysis across time, we can build a full temporal
% network~\cite{masuda2016guide} of communication interactions between people
% in the system.In this study we consider only the network structure and the
% status at that time of the individuals, not the content of messages. 
We study how the email communication is impacted by the hierarchical roles
of individuals. According to the IETF mission statement~\cite{RFC3935}, the
IETF is made up of volunteers who collaborate to develop consensus-based
technical standards. This collaboration should be evident in their
communication practises.

@@

We consider the following questions: 
\begin{enumerate}
  \item
    Are higher levels of the hierarchy growing or becoming more centralised?
    Are they associated with an increased domination of the conversation?
    (\S\ref{sec:org-dyn:hierarchy:rq1})

@@

  \item
    How do people with differing roles communicate with each other?
    Does information flow up or down the hierarchy?
    (\S\ref{sec:org-dyn:hierarchy:rq3})

@@

Question 1 relates to the \emph{steepness} of the hierarchy
\cite{anderson2010functions}, that may exhibit a high amount of centralised
decision making, compared to a more \emph{diffused} \cite{hussain2019voice}
hierarchy where a larger amount of people in higher levels stifle lower
level voice by creating ``bystanders'' who find it hard to speak up. In
general, the conclusions about steep hierarchies are confused, where some
papers conclude a positive and some a negative effect of steep hierarchies.
We explore what this means for the IETF in \S\ref{sec:org-dyn:hierarchy:rq1}.

Questions 2-4 relate to the \emph{power distance}
\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
individuals, which is the number of hierarchy levels between people who are
communicating. The focus is on the impact this has on people's voice or
lack thereof when the distance is large versus small. The conclusions from
this research area usually suggest that even small power distances have a
large effect of reducing the voice of the lower of the two levels. We
tackle this for the IETF in three different ways outlined in 
\S\ref{sec:org-dyn:hierarchy:rq2} -- \S\ref{sec:org-dyn:hierarchy:rq4}.

The voluntary nature of the IETF offers a new perspective on these
questions.  We demonstrate that despite the hierarchy becoming more
diffused, participants in the higher levels in the IETF hierarchy perform a
facilitator role, promoting discussion with those they interact with and
receiving a benefit themselves as a result. We also show that higher level
participants tend to be a focus of discussion with remarks directed upward
toward them whereas Regular Participants engage more in group-discussion
with less focus on an individual.  This suggests that the IETF does not
conform to the notion that power distance has a negative effect on
communication, nor that a more diffused hierarchy reduces the voice of
lower levels.
